Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} Stagnating State Funding for Higher Education and its Effect on Faculty

### 2024-09-27 text/sections/00-abstract.tex
Adjusted formatting of some figures, and changed some bad punctuation.
This is all done.

Public universities employ more lecturers and fewer professors than at any other point in the last thirty years, relative to student enrolment.
At the same time, state funding for higher education has stagnated.
This paper shows that the decline in state funding led to a substitution away from professors toward lecturers at US public universities.
Using a shift-share approach to instrument for state funding, I find that universities employ 4.4\% more lecturers per student following a 10\% funding cut.
This shift is accompanied by a reduction in assistant and full professors by 1.4\% and 1.2\%  per student, respectively.
Incumbent professors' salaries, promotion rates, and quit rates at Illinois universities remain unaffected by funding cuts, indicating that the substitution arose from limiting the hiring of new tenure-track/tenured professors.
Stagnating state funding impacts public universities and faculty, likely contributing to deteriorating student outcomes at public universities seen since the 1990s.

>>>>

Public universities employ more lecturers and fewer professors today than at any other point in the last thirty years, relative to student enrolment.
At the same time, state funding for higher education has stagnated.
This paper shows that the decline in state funding led to a substitution away from professors toward lecturers at US public universities.
Using a shift-share approach to instrument for state funding, I find that universities employ 4.4\% more lecturers per student following a 10\% funding cut.
This shift is accompanied by a reduction in assistant and full professors by 1.4\% and 1.2\%  per student.
These effects are concentrated to 1990--2009, with waning national trends for 2010 and onwards.
Incumbent professors' salaries, promotion rates, and quit rates at Illinois universities remain unaffected by large funding cuts in the 2010s, indicating that the substitution arose from limiting the hiring of new professors.
Stagnating state funding impacts public universities and faculty, likely contributing to deteriorating student outcomes at public universities since the 1990s.

### 2024-09-27 text/sections/01-introduction.tex
This is all done.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\section{Introduction}
\label{sec:intro}
States fund their public universities with roughly the same amount today as they did in the 1990s, despite enrolment rising by over 50\%.
At the same time, public universities have systematically substituted away from professors towards lecturers, who often work on contingent and lower-salaried contracts.
Previous research shows state funding cuts led to worsening student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence on whether these funding cuts impacted faculty.
In this paper I show that falling state funding caused a change in the composition of faculty at US public universities, substituting professors for more lecturers.

Per-student state funding for higher education has stagnated the last thirty years.
The average public university received around \$11,600 of state funding per-student in 1990 and \$8,300 in 2021, while the number of lecturers per-student increased by 109\% and professors per-student decreased by $23$\% over 1990--2021.
I use a shift-share instrumental variables (IV) approach to estimate how faculty are affected by these funding cuts.
Universities employ $4.4$\% more lecturers per-student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer tenure-track professors and $1.2$\% fewer tenured professors per-student.
In a secondary analysis of all public university faculty in the state of Illinois, I show that incumbent faculty were unaffected, in terms of wages, promotion and quit rates.
This implies that the faculty substitution arose by disrupting hiring of new tenure-track/tenured professors, which is supported with supplementary data on professor hiring at US public universities over 2010--2021.
This paper shows the institutional effects of stagnating state funding for higher education, and offers one mechanism for how funding cuts affected student outcomes.

@@

It is a priori unclear how faculty may be affected, as there are various margins that universities may to respond to funding cuts.
On one hand, faculty may be negatively affected by universities cutting salaries or laying-off incumbent faculty.
This happened in 2010 when the University of Illinois System furloughed university staff following budget cuts after the Great Recession \citep{furlough2010}.
Similarly, faculty hiring committees are often cancelled or disrupted in response to funding cuts, meaning there are fewer faculty entering a department to replace those leaving.
On the other hand, lecturers could be hired to fill in teaching duties, on shorter and cheaper contracts with fewer employment protections.
But it is not certain that universities would hire more lecturers in the face of budget cuts.
Public university administrators, motivated by increased competition \citep{hoxby2009changing,urquiola2020markets}, may notice that their rankings are negatively affected by falling professor-lecturer ratios, and so limit departments from hiring more lecturers to replace leaving/retiring professors.
Administrators could even start courting private funding or increase tuition prices to fill funding gaps \citep{bound2019public}, possibly leaving faculty unaffected by state funding cuts.
These are some ways that the long-run stagnation in state funding would not guarantee a substitution towards lecturers at US public universities.

State funding for higher education is not decided randomly;
universities that rely more on state funding may also be institutions that find it difficult to attract new faculty thanks to location, or other factors.
This means that correlations between state funding and faculty outcomes do not necessarily represent causal relationships.
As such, it would be na\"ive to think of state higher education funding as independent of public university faculty, so I employ a shift-share IV approach to identify causal effects of state funding cuts.
The shift-share instrument identifies exogenous changes in state funding by exploiting differences in universities' reliance on state funding, interacted with state-wide shifts in higher education funding --- following the methods of \cite{NBERw23736,NBERw27885}.
Yearly state government budgets are decided without targetting individual universities for funding cuts, so more reliant universities will be more affected by state-wide funding cuts, so the shift-share approach estimates causal effects among more reliant universities.
Additionally, I estimate effect of state funding on faculty multiple years after the initial funding cuts, using the local projections method.\footnote{
    To the best of my knowledge, this is also the first paper to use the local projections method for a shift-share instrument, and to acknowledge the time-series confounding for shift-share IV models when the outcome is in future time periods.
}
This approach allows me to test whether the effects of funding cuts are persistent or universities recover quickly e.g., by quickly resuming faculty hiring in years after funding cuts.

In my university-level analysis (IPEDS), I find that state funding cuts cause public universities to substitute away from tenured and tenure-track professors towards lecturers.
A funding cut of 10\% per-student leads to a fall of 1.4\% in the number of assistant professors per-student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per-student.
Local projection estimates show that these effects linger for up to ten years after the initial funding cut, showing that effects are persistent and not transitory.
Over the same time period, state funding per-student fell by around 35\%, professors per-student fell by 9\%, and lecturers per-student increased by 99\%.
These results show that falls in state funding explain around 53\% of the fall in professors per-student,
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per-student.
% ie., (35 * 0.437) / 99
In my analysis of all Illinois faculty (IBHED), I find that incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
This implies that faculty substitution arose by limiting the hiring of new professors, and is supported by suggestive evidence on professor hiring counts.

I provide the first causal evidence on the impact of state funding on faculty composition and wages.
In a contemporaneous working paper, \cite{hinrichs2022state} use the same national-level variation in university funding to analyse effects on faculty and wages, coming to different conclusions.\footnote{
    An anonymous reviewer made me aware of a similar and contemporary working paper, \cite{hinrichs2022state}; this work analyses a shorter time-period than this paper, and only focuses on national variation.
    \autoref{sec:recession-analysis} replicates their key findings, and shows how their focusing on only the 2012--2019 years explains differences in conclusions.
}
The closest related research on higher education funding examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,webber2017state}.
For faculty outcomes, descriptive work measures how US faculty salaries are correlated with other factors \citep{hilmer2020labor}, and applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
Other papers measure university decision--making after endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects of state funding cuts on faculty composition or salaries, or other outcomes related to university instruction.
My results provide evidence that the long term trends in higher education funding and faculty outcomes are causally related, contributing to the literature on faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing,ehrenberg2012american}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\section{Introduction}
\label{sec:intro}
States fund their public universities with roughly the same amount today as they did in the 1990s, despite enrolment rising by over 50\%.
At the same time, public universities have systematically substituted away from professors towards lecturers, who often work on contingent and lower-salaried contracts.
Previous research shows state funding cuts led to worsening student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence regarding whether these funding cuts impacted faculty.
In this paper I show that falling state funding caused a change in the composition of faculty at US public universities, substituting professors for more lecturers.

Per-student state funding for higher education has stagnated the last thirty years.
The average public university received around \$11,600 of state funding per-student in 1990 and \$8,300 in 2021, while the number of lecturers per-student increased by 109\% and professors per-student decreased by $23$\% over 1990--2021.
I use a shift-share instrumental variables (IV) approach to estimate how faculty are affected by these funding cuts.
Universities employ $4.4$\% more lecturers per-student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer tenure-track professors and $1.2$\% fewer professors per-student.
In a secondary analysis of all public university faculty in the state of Illinois, I show that incumbent faculty were unaffected, in terms of wages, promotion and quit rates.
This implies that the faculty substitution occurred due to disruptions in hiring new professors, which is supported with supplementary data on professor hiring at US public universities over 2010--2021.
This paper shows the institutional effects of stagnating state funding for higher education, and offers one mechanism for how funding cuts affected student outcomes.

@@

It is a priori unclear how faculty may be affected, as there are various margins that universities may respond to funding cuts.
On one hand, faculty may be negatively affected by universities cutting salaries or laying-off incumbent faculty.
This occurred in 2010 when the University of Illinois System furloughed staff in response to budget cuts following the Great Recession \citep{furlough2010}.
Similarly, faculty hiring committees are often cancelled or disrupted in response to funding cuts, meaning there are fewer faculty entering a department to replace those leaving.
On the other hand, lecturers could be hired to fill in teaching duties, on shorter and cheaper contracts with fewer employment protections.
But it is not certain that universities would hire more lecturers in the face of budget cuts.
Public university administrators, motivated by increased competition \citep{hoxby2009changing,urquiola2020markets}, may notice that their rankings are negatively affected by falling professor-lecturer ratios, and so limit departments from hiring more lecturers to replace leaving/retiring professors.
Administrators could even start courting private funding or increase tuition prices to fill funding gaps \citep{bound2019public}, possibly leaving faculty unaffected by state funding cuts.
These are some ways that the long-run stagnation in state funding would not guarantee a substitution towards lecturers at US public universities.

State funding for higher education is not decided randomly;
universities that rely more on state funding may also be institutions that find it difficult to attract new faculty thanks to location, or other factors.
This means that correlations between state funding and faculty outcomes do not necessarily represent causal relationships.
As such, it would be na\"ive to think of state higher education funding as independent of public university faculty, so I employ a shift-share IV approach to identify causal effects of state funding cuts.
The shift-share instrument identifies exogenous changes in state funding by exploiting differences in universities' reliance on state funding, interacted with state-wide shifts in higher education funding --- following the methods of \cite{NBERw23736,NBERw27885}.
Yearly state government budgets are decided without targetting individual universities for funding cuts, so more reliant universities will be more affected by state-wide funding cuts, so the shift-share approach estimates causal effects among universities that are more reliant on state funding.
Additionally, I estimate effect of state funding on faculty multiple years after the initial funding cuts, using the local projections method.\footnote{
    To the best of my knowledge, this is also the first paper to use the local projections method for a shift-share instrument, and to acknowledge the time-series confounding for shift-share IV models when the outcome is in future time periods.
}
This approach allows me to test whether the effects of funding cuts are persistent or universities recover quickly e.g., by quickly resuming faculty hiring in years after funding cuts.

In my university-level analysis (IPEDS), I find that state funding cuts cause public universities to substitute away from tenured and tenure-track professors towards lecturers.
A funding cut of 10\% per-student leads to a fall of 1.4\% in the number of assistant professors per-student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per-student.
Local projection estimates show that these effects linger for up to ten years after the initial funding cut, showing that effects are persistent and not transitory.
Over the same time period, state funding per-student fell by around 35\%, professors per-student fell by 9\%, and lecturers per-student increased by 99\%.
These results show that falls in state funding explain around 53\% of the fall in professors per-student,
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per-student.
% ie., (35 * 0.437) / 99
These effects are concentrated in years before 2010, with the 2010--2021 period showing no evidence for further trends in substitution.  
On the other hand, there are large funding cuts in the 2010s in the state of Illinois.
In my analysis of all Illinois faculty (IBHED), I find that incumbent professors were not meaningfully affected, in terms of total salary, promotion rates, or rates of leaving the Illinois public university system.
This implies that faculty substitution arose by limiting the hiring of new professors, and is supported by suggestive evidence on professor hiring counts.

I provide the first causal evidence on the impact of state funding on faculty composition and wages.
In a contemporaneous working paper, \cite{hinrichs2022state} use the same national-level variation in university funding to analyse effects on faculty and wages, coming to different conclusions.\footnote{
    An anonymous reviewer made me aware of a similar and contemporary working paper, \cite{hinrichs2022state}; this work analyses a shorter time-period than this paper, and only focuses on national variation.
    \autoref{sec:recession-analysis} replicates their key findings, and shows how their focusing on only the 2012--2019 years explains differences in conclusions.
}
The closest related research on higher education funding examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,webber2017state}.
For faculty outcomes, descriptive work measures how US faculty salaries are correlated with other factors \citep{hilmer2020labor}, and applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
Other papers measure university decision--making after endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects of state funding cuts on faculty composition or salaries, or other outcomes related to university instruction.
My results provide evidence that the long-term trends in higher education funding and faculty outcomes are causally related, contributing to the literature on faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing,ehrenberg2012american}.

### 2024-09-27 text/sections/02-data.tex
This is all done.

    \cite{hinrichs2022state} bases the shift--share instrument in the year 2011, despite major upheavals in the level of state funding for state funding for higher education around the 2008 Great Recession, and years following (see \autoref{sec:trends});

@@

    These two major differences explain the different conclusions that our contemporary papers come to; in \autoref{sec:recession-analysis} I replicate the \cite{hinrichs2022state} approach with my primary specification, and further discuses the differences in findings.

@@

This gives a resulting panel data-set, where each row represents a university-year, and includes columns for university funding and tuition revenue, plus total count for each faculty position (lecturer, assistant professor, tenured professor, and total faculty), and other university characteristics.
\autoref{tab:ipeds-summary} presents summary statistics for these variables in IPEDS data.

IPEDS provides information at the university level, but only provides aggregated information within a university.
To investigate outcomes for individual faculty members, I integrate individual-level data for all faculty members at all Illinois public universities 2010--2021.
These data allow me to investigate how individual faculty members are affected by funding cuts to their university, investigating individual-level professor outcomes unavailable in IPEDS data.
IBHED freely provides this information, as the state of Illinois is required to publicly report base salary and benefits\footnote{
    Real salary is computed by scaling nominal salary and benefits to 2021 dollars by the CPI-U.
} for all administrators, faculty members, and instructors employed by each public college or university.\footnote{
    Public Act 96-0266, effective 1 January 2010, is the relevant Illinois law that requires publicly publishing salary data for all public university faculty salary and benefits \citep{illinois-public-act}.
    A pdf copy of Public Act 096-0266 is included in the supplmentary data of this paper.
    This law applies to all nine Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
These data provide the basis to build a panel of Illinois public university faculty across years 2010--2021; I define a faculty member as an individual by their first plus last name and university pairing,\footnote{
    Some faculty are listed multiple times in these panel data.
    For example, Professor Alberto Agustin Lopez-Scala of the University of Illinois Chicago has two appointments in each year 2013--2021, one as an adjunct faculty and one as a department director.
    For this observation I take the highest paid position (department director) as the primary appointment and drop the secondary appointment (adjunct faculty).
    In analysis of faculty exit rate, I collapse the appointments into one and only consider faculty exits as having a faculty appointment versus no appointment.
}
and link this database to IPEDS for data on their university employer.
The Illinois sample represents 16,932 professors in the year 2010 and 15,352 in the year 2021, with summary statistics presented in \autoref{tab:illinois-summary}.
Analysis of professors' in their first year on the job focuses on the subset of professors with observed year of hiring 2011--2021 (i.e., after the panel's first year), representing 1,778 professors in 2011, and 9,099 in 2021.

@@

There are large differences in the average number of professors per student between the private and public sector.
\autoref{fig:fte-perprof} shows that private universities start with an average of 40 students for every one full professor, and little change thereafter.
Public universities start with 40 students per professor, and by 2021 there are 52.4 students professors for every full professor --- with the largest rise coming in the 2008--2011 time period.
The general trend is similar for the number of assistant professors at both public and private universities.
Private and public universities have similar numbers of associate and full professors before the year 2000.
After 2000, the number of professors at public univeristies fell drastically, with the average public university has 6 fewer full professors per hundred students than the average private university by 2021.
Over the same time period, we see the rise in use of lecturers; lecturers were employed at similar rates in private and public sectors in 1990.
Since 1990, both sectors began employing more lecturers on a per-student basis.
In sum, this means public universities have begun relying on lecturers, because they decreased their professor counts, compared to private unviersities who started employing more lecturers and professors.

@@

There was not only a stagnation in state funding in this time period, but also large annual rises and falls, particularly around 2016.
In the calendar year 2015, partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 2015 to June 2016, so is the same as the academic year definition.
}
While most public universities were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring funding to state institutions for 2018.
This means that Illinois public universities exhibit sizable changes in their state funding over 2010--2021, of similar order to those for the rest of the country over 1990--2021.
Additionally, the 2016 episode stemmed entirely from political disagreements, and not from state decisions regarding higher education and its finances, exhibiting how state-level changes in funding affect public universities thanks to unrelated issues \citep{young2020squandered}.

>>>>

    \cite{hinrichs2022state} analyses the years 2012--2019, despite national trends in state funding and faculty counts happening largely in the preceding decades (see \autoref{sec:trends,sec:decade-analysis});

@@

    These two major differences explain the different conclusions that our contemporary papers come to; in \autoref{sec:recession-analysis} I replicate the \cite{hinrichs2022state} approach with my primary specification, and show my results replicate theirs for 2012--2019 (null) and are nonetheless strong for the 1990--2010 time period.

@@

This gives a resulting panel data-set, where each row represents a university-year, and includes columns for university funding and tuition revenue, plus total count for each faculty position (lecturers, assistant professors, professors, and total faculty), and other university characteristics.
I take IPEDS reported count of ``Instructional Staff: Instructor, Lecturer, or No Academic Rank'' and refer to this count as lecturers for brevity, IPEDS count of ``Assistant Professors'' as assistant professors, IPEDS count of ``Associate Professors'' and ``Full Professors'' as professors.\footnote{
    Some universities use the role of associate professor interchangeably with professor (or play the same role in practice as a full professor).
    My definition avoids splitting these hairs by summing all ranks above assistant professor.
\autoref{tab:ipeds-summary} presents summary statistics for these variables in IPEDS data.\footnote{
    IPEDS started collecting information on the composition of instructional staff working part and full-time, as well as salaries, from the 2010 survey and onwards.
    This paper focuses on trends that were more stark in the 1990--2009 time period, so does not present systematic evidence on this topic; \cite{hinrichs2022state} considers this time period and finds no relationship.

IPEDS provides information at the university level, but only provides aggregated information within a university.
To investigate outcomes for individual faculty members, I integrate individual-level data for all faculty members at all Illinois public universities 2010--2021.
These data allow me to investigate how individual faculty members are affected by funding cuts to their university, investigating individual-level professor outcomes unavailable in IPEDS data.
IBHED freely provides this information, as the state of Illinois is required to publicly report base salary and benefits\footnote{
    Real salary is computed by scaling nominal salary and benefits to 2021 dollars by the CPI-U.
} for all administrators, faculty members, and instructors employed by each public college or university.\footnote{
    Public Act 96-0266, effective 1 January 2010, is the relevant Illinois law that requires publicly publishing salary data for all public university faculty salary and benefits \citep{illinois-public-act}.
    A copy of Public Act 096-0266 is included in the supplmentary data of this paper.
    This law applies to all nine Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
These data provide the basis to build a panel of Illinois public university faculty across years 2010--2021; I define a faculty member as an individual by their first plus last name and university pairing,\footnote{
    Some faculty are listed multiple times in these panel data.
    For example, Professor Alberto Agustin Lopez-Scala of the University of Illinois Chicago has two appointments in each year 2013--2021, one as an adjunct faculty and one as a department director.
    For this observation I take the highest paid position (department director) as the primary appointment and drop the secondary appointment (adjunct faculty).
    In analysis of faculty exit rate, I collapse the appointments into one and only consider faculty exits as having a faculty appointment versus no appointment.
}
and link this database to IPEDS for data on their university employer.
The Illinois sample represents 16,932 professors in the year 2010 and 15,352 in the year 2021, with summary statistics presented in \autoref{tab:illinois-summary}.
Analysis of professors' in their first year on the job focuses on the subset of professors with observed year of hiring 2011--2021 (i.e., after the panel's first year), representing 1,778 professors in 2011, and 9,099 in 2021.

@@

There are large differences in the average number of professors per student between the private and public sector.
\autoref{fig:fte-perprof} shows that private universities start with an average of 40 students for every one full professor, and little change thereafter.
Public universities start with 40 students per professor, and by 2021 there are 52.4 students professors for every full professor --- with the largest rise coming in the 2008--2011 time period.
The general trend is similar for the number of assistant professors at both public and private universities.
Private and public universities have similar numbers of associate and full professors before the year 2000.
After 2000, the number of professors at public universities fell drastically, with the average public university has 6 fewer full professors per hundred students than the average private university by 2021.
Over the same time period, we see the rise in use of lecturers; lecturers were employed at similar rates in private and public sectors in 1990.
Since 1990, both sectors began employing more lecturers on a per-student basis.
In sum, this means public universities have begun relying on lecturers, because they decreased their professor counts, compared to private universities who started employing more lecturers and professors.

@@

There was not only a stagnation in state funding in this time period, but also large annual rises and falls, particularly around 2016.
In the calendar year 2015, partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 2015 to June 2016, so is the same as the academic year definition.
}
While most public universities were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring funding to state institutions for 2018.
This means that Illinois public universities exhibit sizeable changes in their state funding over 2010--2021, of similar order to those for the rest of the country over 1990--2021.
Additionally, the 2016 episode stemmed entirely from political disagreements, and not from state decisions regarding higher education and its finances, exhibiting how state-level changes in funding affect public universities thanks to unrelated issues \citep{young2020squandered}.

While the stagnation in state funding was more intense for the national sample of universities over 1990--2009, Illinois in particular experienced larger disruption and state funding cuts for 2010--2021.
Additionally, analysis of individual Illinois professors is only possible as data begin in 2010; while it would be preferred to analyse Illinois professor outcomes for 1990--2009, data availability limits the analysis to 2010--2021.

### 2024-09-27 text/sections/04-empirical.tex
This is all done.

The system exploits the fact that institutions who rely more on state funding will be affected by follow-on state-wide funding cuts.
It is constructed by interacting the average funding for universities in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e., without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
$Z_{i,t}$ is negative to reflect the fact that the long term trend in, and most of the short-run shifts in, state funding for higher education have been negative.\footnote{
    When used in log terms, the instrument is the negative of the logged shock --- i.e., $- \log \left( -Z_{i,t} \right)$ as opposed to log of the negative shock $\log Z_{i,t}$ directly.
}
State funding has been falling, so that the instrument describes shocks to university revenues, mostly in a negative direction.\footnote{
    \label{foot:control}
    \cite{NBERw27885,webber2017state} note the tendency for public universities to respond to state funding cuts by increasing reliance on tuition, where \cite{NBERw23736} specifically instruments for tuition revenues with collected information on legislative tuition price controls.
    It may be argued that tuition revenues are confounder between the causal effect of changes in state funding on a university's total revenues, so that this analysis focuses on state support for higher education (not total revenue), as do \cite{NBERw27885}.
    On the other hand, rises in tuition revenues (per-student) may arise as result of tuition hikes following state funding cuts, which would mean controlling for tuition constitutes a bad control.
    Nonetheless, estimates including tuition revenues (per-student) as a control in the second stage of the IV estimates produce results of very similar magnitude and direction, and so are omitted (see \autoref{sec:results-firststage}).
}

@@

\subsection{Instrumental Variables (IV) Model}
\label{sec:iv-model-uni}
I use the instrument defined in \autoref{eqn:public-instrument} to overcome the endogeneity concerns for state funding to each public university, for an IV approach.
\autoref{eqn:firststage} is the first-stage for exogenous variation in the state funding for university $i$ in year $t$, and \autoref{eqn:secondstage} the second stage for the effect of state funding on faculty outcomes.\footnote{
    The IV approach identifies the local average treatment effect, weighted towards larger values of the instrument in the continuous case.
    So we interpret this treatment effect as a weighted average of causal effects on faculty at public universities to state funding changes, specific to shift-share instruments, among compliers --- i.e., among universities who respond to funding cut and would not have made faculty-outcome changes absent the funding cut.
    Also, we assume that no universities state funding increased in response to the negative shift-share instrument (monotonicity).
}
\begin{equation}
    \label{eqn:secondstage}
    Y_{i,t} = \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{equation}
I estimate the system by two stage least squares, including institution and year fixed effects.\footnote{
    Note that dividing by student count also implicitly controls for the size of the university, so that this model implicitly accounts for yearly variation in professor count and university revenues arising from a university's size of growth/decline.
}
$Y_{i,t}$ represents faculty outcomes for university $i$ in year $t$, $\alpha_i, \gamma_t$ university and year fixed effects, and $\widehat X_{i,t}$ state funding for university $i$ estimated in first-stage \eqref{eqn:firststage}.

@@

\subsection{Effects in Years After the Initial Funding Cut}
\label{sec:local-projections}
Faculty outcomes could be only affected in a transitory manner, where effects are limited to the short-run and fade away quickly, or persistently, where the effects of state funding cuts linger for many years.
I employ a Local Projections (LP) approach to answer the following question: are university faculty affected multiple years after a state funding cut to their university?
The LP method is an empirical model used to estimate dynamic treatment effects when time-series confounding is present, and accommodates an instrument \citep{jorda2005,olea2021inference}.
\begin{align}
    \label{eqn:firststage-lp}
    X_{i,t}   &= \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t} \\
    \label{eqn:secondstage-lp}
    Y_{i,t+k} &= \alpha_i + \gamma_t + \beta_k \widehat X_{i,t} + \varepsilon_{i,t+k}
\end{align}
The approach estimates the first-stage \eqref{eqn:firststage-lp} by ordinary least squares; here this approach is the same as in conventional IV analysis, instrumenting for a university's state funding in year $t$ with state-wide funding shifts in year $t$.
The second-stage \eqref{eqn:secondstage-lp} is estimated by a regression of 
faculty outcomes in future years $t+k$ on (predicted) state funding cuts in year $t$.
Fixed effects, and two-way standard error clustering are maintained throughout, as before.

>>>>

The system exploits the fact that institutions who rely more on state funding will be affected by follow-on state-wide funding cuts.
It is constructed by interacting the average funding for universities in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e., without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
$Z_{i,t}$ is negative to reflect the fact that the long term trend in, and most of the short-run shifts in, state funding for higher education have been negative.\footnote{
    When used in log terms, the instrument is the negative of the logged shock --- i.e., $- \log \left( -Z_{i,t} \right)$ as opposed to log of the negative shock $\log Z_{i,t}$ directly.
}
State funding has been falling, so that the instrument describes shocks to university revenues, mostly in a negative direction.\footnote{
    \label{foot:control}
    \cite{NBERw27885,webber2017state} note the tendency for public universities to respond to state funding cuts by increasing reliance on tuition, where \cite{NBERw23736} specifically instruments for tuition revenues with collected information on legislative tuition price controls.
    It may be argued that tuition revenues are confounder between the causal effect of changes in state funding on a university's total revenues, so that this analysis focuses on state support for higher education (not total revenue), as do \cite{NBERw27885}.
    On the other hand, rises in tuition revenues (per-student) may arise as result of tuition hikes following state funding cuts, which would mean controlling for tuition constitutes a bad control.
    Nonetheless, estimates including tuition revenues (per-student) as a control in the second stage of the IV estimates produce results of very similar magnitude and direction, and so are omitted (see \autoref{sec:results-firststage}).
}
It might be the case that funding cuts have systematically different effects on universities, and faculty, than funding increases.
The identification strategy here identifies the effect of funding changes, and refers to funding cuts throughout to reflect the systematic falls of state funding (per student) at US public universities.
I leave it to future research to develop a separate identification strategy that can distinguish the different effects of funding cuts and increases; this paper can not.

@@

\subsection{Instrumental Variables (IV) Model}
\label{sec:iv-model-uni}
I use the instrument defined in \autoref{eqn:public-instrument} to overcome the endogeneity concerns for state funding to each public university, for an IV approach.
\autoref{eqn:firststage} is the first-stage for exogenous variation in the state funding for university $i$ in year $t$, and \autoref{eqn:secondstage} the second stage for the effect of state funding on faculty outcomes.\footnote{
    The IV approach identifies the local average treatment effect, weighted towards larger values of the instrument in the continuous case.
    So we interpret this treatment effect as a weighted average of causal effects on faculty at public universities to state funding changes, specific to shift-share instruments, among compliers --- i.e., among universities who respond to funding cut and would not have made faculty-outcome changes absent the funding cut.
    Also, we assume that no universities state funding increased in response to the negative shift-share instrument (monotonicity).
}
\begin{equation}
    \label{eqn:secondstage}
    Y_{i,t} = \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{equation}
I estimate the system by two-stage least-squares, including institution and year fixed effects.
$Y_{i,t}$ represents faculty outcomes for university $i$ in year $t$, $\alpha_i, \gamma_t$ university and year fixed effects, and $\widehat X_{i,t}$ state funding for university $i$ estimated in first-stage \eqref{eqn:firststage}.
Dividing by student count in both outcome and explanatory variables also implicitly controls for the size of the university, so that this model implicitly accounts for yearly variation in professor count and university revenues arising from a university's size of growth/decline.
Nonetheless, including institution and year fixed effects also controls for the threat to identification of funding cuts possibly being aimed at larger institutions --- even if the primary specifications did not account for student population sizes.

@@

\subsection{Effects in Years After the Initial Funding Cut}
\label{sec:local-projections}
Faculty outcomes could be only affected in a transitory manner, where effects are limited to the short-run and fade away quickly, or persistently, where the effects of state funding cuts linger for many years.
I employ a Local Projections (LP) approach to answer the following question: are university faculty affected multiple years after a state funding cut to their university?
The LP method is an empirical model used to estimate dynamic treatment effects when time-series confounding is present, and accommodates an instrument \citep{jorda2005,olea2021inference}.
\begin{align}
    \label{eqn:firststage-lp}
    X_{i,t}   &= \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t} \\
    \label{eqn:secondstage-lp}
    Y_{i,t+k} &= \alpha_i + \gamma_t + \beta_k \widehat X_{i,t} + \varepsilon_{i,t+k}
\end{align}
The approach estimates the first-stage \eqref{eqn:firststage-lp} by ordinary least squares; here this approach is the same as in conventional IV analysis, instrumenting for a university's state funding in year $t$ with state-wide funding shifts in year $t$.
The second-stage \eqref{eqn:secondstage-lp} is estimated by a regression of 
faculty outcomes in future years $t+k$ on (predicted) state funding cuts in year $t$.
Fixed effects, and two-way standard error clustering are maintained throughout, as before.
This method requires no further identifying assumptions beyond those for the regular IV methods.
If the conditions for the shift-share IV apply (\autoref{sec:iv-model-uni}), then identification holds for using LP methods.
